---
title: "Wearable Health Data Analysis: Turning Data into Actionable Health Insights"
author: "Dr. R. Craig Stillwell"
output: github_document
---

## Introduction

This report presents a comprehensive data analysis using wearable health data from Apple Watch and Fitbit. The analysis includes data preparation, clustering to identify user patterns, and predictive modeling to classify activity types and predict heart rate. The insights derived from this analysis are aimed at helping users make informed health decisions, improving device algorithms, and supporting the creation of personalized health recommendations.

```{r setup, include=FALSE}
# Load required packages and install if missing
packages <- c("tidyverse", "cluster", "factoextra", "caret", "randomForest", "Metrics", "NbClust")
install.packages(setdiff(packages, rownames(installed.packages())))

library(NbClust)
library(tidyverse)
library(cluster)
library(factoextra)
library(caret)
library(randomForest)
library(Metrics)
```
## Data Preparation

```{r data-cleaning, echo=TRUE}
# Load the dataset
data <- read.csv('data/aw_fb_data.csv')

# Initial data check
str(data)
summary(data)
colSums(is.na(data))

# Data Cleaning: Remove columns with all missing values and drop rows without an activity label
data <- data %>%
  select_if(~ !all(is.na(.))) %>%  # Remove columns with all missing values
  mutate(activity = as.character(activity)) %>%
  drop_na(activity)                # Drop rows with NA in 'activity'

# Save cleaned data
write.csv(data, file = "/Users/stillwell/Documents/R/Apple Watch and Fitbit Health Data Portfolio/data/cleaned_aw_fb_data.csv")
```
## Clustering Analysis

We use K-means clustering to explore distinct patterns in heart rate and step count, identifying user groups with similar health metrics, which may represent different activity and lifestyle patterns.

```{r clustering, echo=TRUE}
# Select relevant columns and normalize data for clustering
clustering_data <- data %>%
  select(hear_rate, steps, resting_heart)
clustering_data_normalized <- scale(clustering_data)

# Determine the optimal number of clusters using silhouette method and save result
silhouette_result <- NbClust(clustering_data_normalized, distance = "euclidean", min.nc = 2, max.nc = 10, method = "kmeans", index = "silhouette")

# Retrieve the best number of clusters
optimal_clusters <- silhouette_result$Best.nc[1]
print(paste("Optimal number of clusters based on silhouette method:", optimal_clusters))

# Visualize with fviz_nbclust to confirm
fviz_nbclust(clustering_data_normalized, kmeans, method = "silhouette") +
  labs(title = "Optimal Number of Clusters (Silhouette Method)")
```
## Optimal Cluster Count

The silhouette method suggests an optimal cluster count of four, indicating distinct user profiles, which could represent different lifestyle patterns (e.g., sedentary or moderately active) and health states.
```{r kmean, echo=TRUE}
# Perform K-means clustering and add results to data
set.seed(123)
kmeans_result <- kmeans(clustering_data_normalized, centers = 4, nstart = 25)
data$cluster <- as.factor(kmeans_result$cluster)

# Visualize clustering results
ggplot(data, aes(x = steps, y = hear_rate, color = cluster, shape = activity)) +
  geom_point(alpha = 0.7, size = 3) +
  theme_minimal(base_size = 15) +
  labs(
    title = "K-Means Clustering of Heart Rate Patterns",
    subtitle = "Identifying Patterns in Wearable Data Using K-Means Clustering",
    x = "Steps",
    y = "Heart Rate"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 12, face = "italic"),
    legend.position = "bottom",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
```
## Predictive Modeling

Model 1: Predicting Heart Rate

Using a Random Forest model to predict heart rate based on wearable data metrics provides insights for both users and developers to understand and anticipate health changes.
```{r predictive-modeling, echo=TRUE}
# Feature selection and train-test split
features <- data %>%
  select(steps, calories, intensity_karvonen, sd_norm_heart, steps_times_distance, activity, hear_rate) %>%
  drop_na()
features$activity <- as.factor(features$activity)
set.seed(123)
trainIndex <- createDataPartition(features$hear_rate, p = 0.8, list = FALSE)
train_data <- features[trainIndex, ]
test_data <- features[-trainIndex, ]

# Train a Random Forest model for heart rate prediction
rf_model_hr <- randomForest(hear_rate ~ ., data = train_data, importance = TRUE)
predictions_hr <- predict(rf_model_hr, newdata = test_data)
rmse_value <- rmse(test_data$hear_rate, predictions_hr)
r2_value <- cor(test_data$hear_rate, predictions_hr)^2

# Print formatted RMSE and R-squared values
rmse_value <- rmse(test_data$hear_rate, predictions_hr)
r2_value <- cor(test_data$hear_rate, predictions_hr)^2

# Print results with formatted output
print(sprintf("Model Metrics:\n- Root Mean Square Error (RMSE): %.2f\n- R-squared (R²): %.3f", rmse_value, r2_value))
```
## Model 1

The model achieved an RMSE of 9.36 and an R² of 0.896, demonstrating that heart rate can be effectively predicted from activity metrics. This information could improve device algorithms for health tracking.

## Predictive Modeling

Model 2: Classifying Activity Type

Using a Random Forest classifier to identify activity types based on health data provides actionable insights into daily activity levels, encouraging users to optimize their health routines.
```{r confusion-matrix, echo=TRUE}
# Train a Random Forest model for activity classification
rf_model_activity <- randomForest(activity ~ ., data = train_data, importance = TRUE)
predictions_activity <- predict(rf_model_activity, newdata = test_data)
conf_matrix_activity <- confusionMatrix(predictions_activity, test_data$activity)
print(conf_matrix_activity)
varImpPlot(rf_model_activity, main = "Feature Importance in Predicting Activity Type")
```
## 
Model 2

The confusion matrix displays correct and incorrect predictions for each activity type, showing high accuracy, especially for high-intensity activities. This classification model helps improve activity tracking and distinguish between similar activity types.

Conclusion

This analysis demonstrates how wearable health data can reveal patterns in user activity and predict key metrics like heart rate. These insights support the development of personalized health recommendations and improve wearable technology, allowing users to make informed health decisions based on data-driven insights.