---
title: "Wearable Health Data Analysis"
author: "Dr. R. Craig Stillwell"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
---

## Introduction

This report presents a comprehensive data analysis using wearable health data from Apple Watch and Fitbit. The analysis includes data preparation, clustering to identify user patterns, and predictive modeling to classify activity types and predict heart rate. The insights derived from this analysis can inform device improvements, user engagement, and personalized health recommendations.

```{r setup, include=FALSE}
# Load required packages and install if missing
if (!requireNamespace("tidyverse")) install.packages("tidyverse")
if (!requireNamespace("cluster")) install.packages("cluster")
if (!requireNamespace("factoextra")) install.packages("factoextra")
if (!requireNamespace("caret")) install.packages("caret")
if (!requireNamespace("randomForest")) install.packages("randomForest")
if (!requireNamespace("Metrics")) install.packages("Metrics")

library(tidyverse)
library(cluster)
library(factoextra)
library(caret)
library(randomForest)
library(Metrics)
```

## Data Preparation

```{r data-cleaning, echo=TRUE}
# Load the dataset
data <- read.csv('data/aw_fb_data.csv')

# Initial data check
str(data)
summary(data)

# Data Cleaning
data <- data %>%
  select_if(~ !all(is.na(.))) %>%  # Remove columns with all missing values
  mutate(activity = as.character(activity)) %>%
  drop_na(activity)                # Drop rows with NA in 'activity'

# Save cleaned data
write.csv(data, file = "/Users/stillwell/Documents/R/Apple Watch and Fitbit Health Data Portfolio/data/cleaned_aw_fb_data.csv")
```

## Clustering Analysis

We use K-means clustering to explore distinct patterns in heart rate and step count, identifying user groups with similar health metrics.

```{r clustering, echo=TRUE}
# Select relevant columns and normalize data for clustering
clustering_data <- data %>%
  select(hear_rate, steps, resting_heart)
clustering_data_normalized <- scale(clustering_data)

# Determine the optimal number of clusters using the silhouette method
fviz_nbclust(clustering_data_normalized, kmeans, method = "silhouette") +
  labs(title = "Optimal Number of Clusters (Silhouette Method)")
```

The silhouette method suggests an optimal cluster count of four, indicating distinct user profiles, which could represent lifestyle patterns (e.g., sedentary or moderately active) and health states.

```{r kmean, echo=TRUE}
# Perform K-means clustering and add results to data
set.seed(123)
kmeans_result <- kmeans(clustering_data_normalized, centers = 4, nstart = 25)
data$cluster <- as.factor(kmeans_result$cluster)

# Visualize clustering results
ggplot(data, aes(x = steps, y = hear_rate, color = cluster, shape = activity)) +
  geom_point(alpha = 0.7, size = 3) +
  theme_minimal(base_size = 15) +
  labs(title = "K-Means Clustering of Heart Rate Patterns",
       subtitle = "Identifying Patterns in Wearable Data Using K-Means Clustering",
       x = "Steps",
       y = "Heart Rate") +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))
```

## Predictive Modeling

### Model 1: Predicting Heart Rate

A Random Forest model predicts heart rate using key metrics such as steps, calories, and intensity.

```{r predictive-modeling, echo=TRUE}
# Feature selection and train-test split
features <- data %>%
  select(steps, calories, intensity_karvonen, sd_norm_heart, steps_times_distance, activity, hear_rate) %>%
  drop_na()
features$activity <- as.factor(features$activity)
set.seed(123)
trainIndex <- createDataPartition(features$hear_rate, p = 0.8, list = FALSE)
train_data <- features[trainIndex, ]
test_data <- features[-trainIndex, ]

# Train a Random Forest model for heart rate prediction
rf_model_hr <- randomForest(hear_rate ~ ., data = train_data, importance = TRUE)
predictions_hr <- predict(rf_model_hr, newdata = test_data)
rmse_value <- rmse(test_data$hear_rate, predictions_hr)
r2_value <- cor(test_data$hear_rate, predictions_hr)^2

# Print results
cat("Root Mean Square Error (RMSE):", rmse_value, "\n")
cat("R-squared:", r2_value, "\n")
varImpPlot(rf_model_hr, main = "Feature Importance in Predicting Heart Rate")
```

The model achieved an RMSE of 9.362 and an R² of 0.8959, demonstrating that heart rate can be effectively predicted from activity metrics. This information could improve device algorithms for health tracking.

### Model 2: Classifying Activity Type

We use a Random Forest classifier to identify activity types, offering real-time feedback on workout intensity.

```{r confusion-matrix, echo=TRUE}
# Train a Random Forest model for activity classification
rf_model_activity <- randomForest(activity ~ ., data = train_data, importance = TRUE)
predictions_activity <- predict(rf_model_activity, newdata = test_data)
conf_matrix_activity <- confusionMatrix(predictions_activity, test_data$activity)
print(conf_matrix_activity)
varImpPlot(rf_model_activity, main = "Feature Importance in Predicting Activity Type")
```

The confusion matrix displays correct and incorrect predictions for each activity type, showing high accuracy, especially for high-intensity activities. This classification model helps improve activity tracking and distinguish between similar activity types.

## Conclusion

This analysis demonstrates how wearable health data can reveal patterns in user activity and predict key metrics like heart rate. These insights can support the development of personalized health recommendations and improve wearable technology.